\chapter{Analysis modules\label{cha:Analysis-modules}}

\section{Generating grids of model results (module GRI)\ref{sec:gri-module}}

Module GRI provides a facility for the construction of grids (e.g.
for contour plotting) in a format compatible with SURFER\texttrademark,
Matlab\texttrademark, or other software packages. Within the grid
module, the window to be gridded must be specified, along with the
number of points on the long axis of the grid. 

5.7.1 Selecting the output file type (opt directive) The opt directive
instructs the grid module which type of output file to create. Usage:

opt grid-type The opt directive expects one parameter, as follows.
grid-type Choose surfer for an ASCII SURFER\texttrademark-compatible grid (with
the extension .grd) and matlab for

an ASCII MATLABTM-compatible grid (with the extension .m). If the
OPT directive is omitted, the grid-type will default to surfer.

Example To select a MatlabTM-compatible output file, issue the directive

opt matlab

5.7.2 Defining the grid window (win directive) Defines the window
to be gridded. Usage :

win (x1,y1) (x2,y2) The win directive expects the following parameters.
(x1,y1) The lower-left corner of the (rectangular) region to be gridded.
(x2,y2) The upper-right corner of the (rectangular) region to be gridded.

Example

win (-100.0,-100.0) (100.0,100.0)

sets the lower-left and upper-right corners of the window for the
GRI module at the coordinates \textbackslash{}Gamma \textbackslash{}Delta
\textbackslash{}Gamma 100\textbackslash{}Delta \textbackslash{}Delta
\textbackslash{}Gamma 100\textbackslash{}Theta and\textbackslash{}Gamma 

100\textbackslash{}Delta 100\textbackslash{}Theta , respectively.

5.7.3 Choosing the grid resolution (dim directive) Sets the number
of grid points along the long axis of the window. Usage:

dim npts The dim directive expects one parameter as follows. npts
The number of evenly-spaced points to compute along the long axis
of the rectangular grid region. Module GRI

will compute an appropriate number of points along the short axis
to ensure that the grid has regular spacing in both directions.

Example To make grid(s) with a resolution of 50 points along the long
axis, issue the directive

gri 50

5.7.4 Computing a grid and writing it to a file (directives hea, pot,
psi, q\_x, and q\_y) Once the grid type, grid region, and grid resolution
are specified (see the directives opt, dim, and win above), grid files
may be computed for a variety of model output values. Currently the
following directives are available:

HEA - Create a grid of heads Usage:

hea base-filename generates a grid of the potentiometric head f on
the file base-filename\_head.grd (or base-filename\_head.m)

POT - Create a grid of potentials Usage:

pot base-filename generates a grid of the discharge potential \textbackslash{}Phi
on the file base-filename\_potential.grd (or base-filename\_potential.m)

PSI - Create a grid of stream functions Usage:

psi base-filename generates a grid of the streamfunction \textbackslash{}Psi
on the file base-filename\_psi.grd (or base-filename\_psi.m)

Q\_X - Create a grid of discharges in the x-direction Usage:

q\_x base-filename generates a grid of the total aquifer discharge
in the x-direction potentiometric heads on the file base-filename\_qx.grd
(or base-filename\_qx.m)

Q\_Y - Create a grid of discharges in the y-direction Usage:

q\_y base-filename generates a grid of the total aquifer discharge
in the y-direction on the file base-filename\_qy.grd (or base-filename\_qy.m)


\section{Data extraction (module INQ)\label{sec:inq-module}}

Module INQ provides a facility for extracting computed values from a ModAEM solution and
writing them to output files. Two output formats are supported:
\begin{description}
\item[{Traditional (\texttt{opt inq})}] All output is written to a single file opened by the
  \textsf{fil} directive. This is the default.
\item[{CSV (\texttt{opt csv})}] Each directive type writes to its own comma-separated value
  file, suitable for direct import into Python, R, Excel, or other tools.
\end{description}
The INQ module is entered with the \textsf{inq} directive and terminated with \textsf{end}.

\subsection{Selecting the output format (\texttt{opt} directive)}

\begin{verbatim}
opt inq | csv
\end{verbatim}

\texttt{opt inq} writes all output to the single file named by \textsf{fil} (default behaviour).
\texttt{opt csv} writes each directive type to a separate CSV file. The \textsf{fil} directive
sets a file-name prefix; if omitted, files are named \texttt{<type>\_inq.csv}.

\subsection{Naming the output file (\texttt{fil} directive)}

\begin{verbatim}
fil filename
\end{verbatim}

In \emph{traditional mode}, \textsf{fil} opens \texttt{filename} for writing and all subsequent
extraction output is written to it.

In \emph{CSV mode}, \textsf{fil} sets a prefix for all CSV file names. For example,
\texttt{fil mymodel} causes extraction directives to write to files named
\texttt{mymodel\_hea.csv}, \texttt{mymodel\_ls1.csv}, etc. If \textsf{fil} is omitted in CSV
mode, files are named \texttt{hea\_inq.csv}, \texttt{ls1\_inq.csv}, etc.

\subsection{Spatial extraction directives}

These directives evaluate the solution at one or more points or paths. In traditional mode all
output goes to the file opened by \textsf{fil}. In CSV mode each directive type appends to its
own CSV file; a header row is written the first time a given type is used within an INQ block.
Multiple occurrences of the same directive type accumulate rows in the same file.

\begin{description}
\item[\textsf{hea} $(x,y)$ label] Computes the potentiometric head at $(x,y)$. CSV columns:
  \texttt{label, x, y, head}.
\item[\textsf{dis} $(x,y)$ label] Computes the depth-integrated discharge vector. CSV columns:
  \texttt{label, x, y, discharge\_x, discharge\_y}.
\item[\textsf{vel} $(x,y)$ label] Computes the seepage velocity vector. CSV columns:
  \texttt{label, x, y, velocity\_x, velocity\_y}.
\item[\textsf{pot} $(x,y)$ label] Computes the discharge potential. CSV columns:
  \texttt{label, x, y, potential\_r, potential\_i}.
\item[\textsf{flo} $(x_1,y_1)$ $(x_2,y_2)$ label] Computes the total volumetric flow crossing
  the segment. CSV columns: \texttt{label, x1, y1, x2, y2, flow}.
\item[\textsf{bdy} $(x_1,y_1)$ $(x_2,y_2)$ label] Computes the head at the midpoint and the
  integrated flow crossing the segment. CSV columns: \texttt{label, x1, y1, x2, y2, head, flux}.
\item[\textsf{rch} $(x,y)$ label] Computes the net recharge rate. CSV columns:
  \texttt{label, x, y, recharge}.
\item[\textsf{sat} $(x,y)$ label] Computes the saturated thickness. CSV columns:
  \texttt{label, x, y, sat\_thickness}.
\item[\textsf{gag} gage-id label] Returns the routed flow at the downstream end of an LS2 stream
  string. CSV columns: \texttt{gage\_id, label, flow}.
\end{description}

\subsection{Module inquiry directives}

These directives write a table of element-level check values for a given module. In traditional
mode output goes to the file opened by \textsf{fil}. In CSV mode each directive writes a
separate file with a \texttt{tag} column as the first field identifying the record type.

\begin{description}
\item[\textsf{wl0} / \textsf{wl1}] Specified-discharge / specified-head wells.
  CSV files: \texttt{wl0.csv}, \texttt{wl1.csv}.
\item[\textsf{ls0} / \textsf{ls1} / \textsf{ls2}] Line-sink packages.
  CSV files: \texttt{ls0.csv}, \texttt{ls1.csv}, \texttt{ls2.csv}.
\item[\textsf{hb0}] No-flow barrier strings. CSV file: \texttt{hb0.csv}.
\item[\textsf{as0}] Area-sink package. CSV file: \texttt{as0.csv}.
\item[\textsf{aqu}] Aquifer reference and boundary elements. Writes up to three sections:
  reference point (\texttt{AQU} records), boundary elements (\texttt{BDY} records), and
  inhomogeneity interface elements (\texttt{IN0} records). In CSV mode the sections are
  separated by a blank line, each with its own header row. CSV file: \texttt{aqu.csv}.
\item[\textsf{cw0}] Collector wells. Writes two sections: well summary (\texttt{CW0} records)
  and radial arm elements (\texttt{CWR} records). In CSV mode the sections are separated by a
  blank line. CSV file: \texttt{cw0.csv}.
\end{description}

\subsection{Examples}

Traditional mode:
\begin{verbatim}
inq
  fil mymodel.inq
  hea (0.0,0.0) 1
  hea (100.0,0.0) 2
  flo (-50.0,-100.0) (-50.0,100.0) 10
  ls1
  aqu
end
\end{verbatim}

CSV mode (writes \texttt{mymodel\_hea.csv}, \texttt{mymodel\_flo.csv},
\texttt{mymodel\_ls1.csv}, \texttt{mymodel\_aqu.csv}):
\begin{verbatim}
inq
  opt csv
  fil mymodel
  hea (0.0,0.0) 1
  hea (100.0,0.0) 2
  flo (-50.0,-100.0) (-50.0,100.0) 10
  ls1
  aqu
end
\end{verbatim}

\section{Streamline tracing (module TR0)\label{sec:tr0-module}}

5.8 Module TR0 - Trace The TR0 directive instructs ModAEM to enter
the trace module, which is used to trace 2-D streamlines. The TR0
directive must have a matching END directive. Within the TR0 module,
the following directives are valid:

WIN - Set the tracing window. Default tuning parameters are derived
from the window size.

TUN - Set tuning parameters Sets tuning parameters for the tracing
algorithm. Usage:

TUN step prox frac small step The base step size prox The proximity
(in terms of the current step size) to boundary conditions for reducing
the step size frac The factor for step size reductions small Smallest
allowable step size

TIM - Specify maximum time allowed for particle tracing POI - Release
a single particle at the specified location LIN - Release particles
along a line N particles along a line

GRI - Release a grid of particles in the sub-window WL0 - Release
N particles in reverse from the well bore of a WL0 (discharge-specified
well) element. WL1 - Release N particles in reverse from the well
bore of a WL1 (head-specified well) element.
